---
title: "IBS spatial phylogenetics workshop"
author: "Matthew Kling"
date: "2022-09-26"
output: html_document
editor_options: 
  chunk_output_type: inline
---

# SETUP

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

# install.packages(
#       "ape", "phyloregion", "canaper",            # phylo libraries
#       "terra", "geosphere", "prioritizr", "gdm",  # spatial libraries
#       "pals", "vegan", "tidyverse", "Rsymphony")  # other utilities

# install.packages("canaper", repos = "https://ropensci.r-universe.dev")

library(ape)
library(tidyverse)
library(terra)
library(phyloregion)
library(canaper)
library(prioritizr)
library(pals)
library(vegan)
library(geosphere)

source("../../code/R/functions.r")

select <- dplyr::select
```

------------------------------------------------------------------------

# WARMUP

## Phylogenies

Let's start with a quick primer on phylogenetic trees in R. We'll use `ape`, the core R phylogenetics library. (Other libraries to be aware of for working with phylogenies include `phytools`, `picante`, and `ggtree`, among others.)

```{r phylo}

# load a tree and plot it
tree <- read.tree("../../data/mishler_2014/mishler_2014_acacia.tre")
plot(tree, type = "fan", show.tip.label = F)

# explore the data structure of a phylogeny


# plot the branches connecting a set of tips (e.g. a local community).
# PD is the lengths of these branches, plus the edges connecting them to the root.
community <- sample(tree$tip.label, 5)
phyplot(tree, connect = community, 
        show.tip.label = F, type = "fan")
```

#### Exercises:

-   Load and plot the mammal phylogeny (Upham et al. 2019).
-   Plot the subtree representing the marsupials. The most recent common ancestor of the two species listed below is the crown node of all marsupials; you can find it using `getMRCA()` and prune it from the full tree using `extract.clade()`.
-   Plot the full mammal tree with marsupials highlighted in red. You can use `phytools::getDescendants()` to index all the descendants of the marsupial MRCA.

```{r phylo_ex}
marsupials <- c("Lestoros_inca", "Lagostrophus_fasciatus")
```

------------------------------------------------------------------------

## Spatial data in R

For a spatial phylogenetic analysis, we need to define geographic areas of interest, within which organisms are considered part of the same community. These areas can be grid cells or can be irregular polygons, but we'll focus on the former here as the latter are conceptually similar but more complex to work with.

We'll be working with raster data and point data. The raster data comprise a spatial grid of presences and absences. Point data represent occurrence localities, which we'll convert to grids of presences and absences before analysis.

Here's a quick demo of loading and visualizing some spatial data.

```{r spatial}

# raster data, read in with the terra package
r <- rast("../../data/jetz_2012/bbs_occ.tif")
# r
# names(r)
plot(r[[1:4]], col = "green")
plot(subset(r, "Accipiter cooperii"))
plot(sum(r, na.rm = T))

# point data, read in as a data frame
p <- read_csv("../../data/mishler_2014/mishler_2014_acacia_points.csv")
# glimpse(p)
ggplot(p, aes(Longitude, Latitude)) + 
      geom_point()

# we can "rasterize" points by rounding the coordinates
# (there are fancier approaches but we'll keep it simple here)
pr <- p %>% mutate(x = round(Longitude), y = round(Latitude))
ggplot(pr, aes(x, y)) + 
      geom_raster()

# taking this a step farther, lets' convert our rounded coordinates 
# for one species into to a proper raster object
rp <- pr %>%
      filter(taxa == taxa[1000]) %>%
      select(x, y) %>%
      distinct() %>%
      mutate(present = T) %>%
      rast()
plot(rp, col = "black")
```

#### Exercises:

-   Load the terrestrial mammal spatial data using `rast()`
-   Plot a map of the range of the gray wolf (Canis lupus).
-   PLot a map of canid species richness (all species in the genus Canis).
-   Plot a map of overall mammal species richness.

```{r spatial_ex}

```

------------------------------------------------------------------------

# SPATIAL PHYLOGENETIC DATA

At a minimum, every spatial phylogenetic dataset consists of a phylogeny and a spatial dataset representing occurrence locations of the terminal taxa. We'll use four different datasets as examples.

To keep things tidy, let's start by loading the data for each one, and getting the pieces into a standardized format. We'll package each dataset in a list with three pieces:

-   tree: the phylogeny
-   comm: a community matrix, with a row for each site and a column for each terminal taxon
-   xy: a data frame storing the spatial coordinates of each row of the community matrix

To avoid issues with some of the algorithms used later on, we need to do a few things to clean up the datasets, including removing taxa that aren't present in both the spatial and phylogenetic datasets, and removing unoccupied sites from the community matrix. To do this we'll use a function called `clean_dataset()` defined in `functions.R`.


#### Calfiornia vascular plants (from [Thornhill et al. 2017](https://doi.org/10.1186/s12915-017-0435-x)):

```{r flora_data}

# phylogeny
tree <- read.nexus("../../data/thornhill_2017/thornhill_2017_chronogram.nex")

# spatial data
occ <- read_csv("../../data/thornhill_2017/thornhill_2017_occ.csv")

# # compare species lists
# all.equal(sort(tree$tip.label),
#           sort(unique(occ$otu)))

# summarize to lat-long grid
rnd <- function(x, y) round(x/y)*y # round x to the nearest y
occ <- occ %>%
      mutate(x = rnd(x, .5),
             y = rnd(y, .5)) %>%
      distinct()

# construct community matrix
comm <- occ %>%
      mutate(pres = 1) %>%
      spread(otu, pres, fill = 0)
xy <- select(comm, x, y)
comm <- select(comm, -x, -y) %>%
      as.matrix()

# package data
flora <- list(tree = tree,
              comm = comm,
              xy = xy) %>%
      clean_dataset()
```

#### Australian acacia (from [Mishler et al. 2014](https://doi.org/10.1038/ncomms5473)):

```{r acacia_data}

# phylogeny
tree <- read.tree("../../data/mishler_2014/mishler_2014_acacia.tre")

# spatial data
occ <- read_csv("../../data/mishler_2014/mishler_2014_acacia_points.csv")

# construct community matrix
occ <- occ %>%
      select(otu = taxa, x = Longitude, y = Latitude) %>%
      mutate(x = rnd(x, 1), # summarize to 0.5 degree lat-lon grid
             y = rnd(y, 1)) %>%
      distinct()
comm <- occ %>%
      mutate(pres = 1) %>%
      spread(otu, pres, fill = 0)
xy <- select(comm, x, y)
comm <- select(comm, -x, -y) %>%
      as.matrix()

# package data
acacia <- list(tree = tree,
               comm = comm,
               xy = xy) %>%
      clean_dataset()
```

#### American birds (phylogeny from [Jetz et al. 2012](https://doi.org/10.1038/nature11631), spatial data from [US Breeding Bird Survey](https://www.pwrc.usgs.gov/bbs/rawdata/)):

```{r bird_data}

# phylogeny (one of 10,000 posterior trees from Jetz 2012)
tree <- read.nexus("../../data/jetz_2012/jetz_2012_tree0001.nex")

# spatial data
r <- rast("../../data/jetz_2012/bbs_occ.tif")

# construct community matrix
comm <- values(r)

# spatial coordinates
xy <- crds(r[[1]], df = T, na.rm = F)

# package data
birds <- list(tree = tree,
              comm = comm,
              xy = xy) %>%
      clean_dataset()
```

#### Mammals (phylogeny from [Upham et al. 2019](https://doi.org/10.1371/journal.pbio.3000494), spatial data from [IUCN](https://www.iucnredlist.org/resources/spatial-data-download)):

```{r mammal_data}

# phylogeny (one of 10,000 posterior trees from Upham 2019)
tree <- read.tree("../../data/upham_2019/upham_2019_tree0000.tre")

# spatial data
r <- rast("../../data/upham_2019/iucn_mammals.tif")

# construct community matrix
comm <- values(r)

# spatial coordinates
xy <- crds(r[[1]], df = T, na.rm = F)

# package data
mammals <- list(tree = tree,
                comm = comm,
                xy = xy) %>%
      clean_dataset()
```


#### Exercises:

-   Plot the phylogeny for each of the datasets.
-   Plot the spatial coordinates for each of the datasets.
-   Identify the number of taxa and number of sites in each dataset.

```{r data_ex}

```


------------------------------------------------------------------------

# ALPHA DIVERSITY

Alpha phylogenetic diversity (and related metrics like phylogenetic endemism) represent the amount of evolutionary history present in a single community. We'll look at functions from some existing packages that can calculate phylodiversity metrics, and reproduce these with some of our own calculations. We'll also cover methods for significance testing, and explore the sensitivity of these tests to spatial scale. And we'll take a look at how alternative evolutionary branch lengths influence results.


#### Pre-packaged functions

```{r alpha}

ds <- birds

# with phyloregion library
scomm <- ds$comm %>% Matrix(sparse = TRUE)
pr <- ds$xy %>%
      mutate(pd = PD(scomm, ds$tree),
             pe = phylo_endemism(scomm, ds$tree),
             we = weighted_endemism(scomm))
pr %>% carto("pd")

# with canaper library
cpr <- cpr_rand_test(ds$comm, ds$tree, 
                     null_model = "curveball",
                     n_reps = 1) %>%
      bind_cols(ds$xy)
cpr %>% carto("pd_obs")
cpr %>% carto("pe_obs")

# plot a local community on the phylogeny
par(mfrow = c(1, 2))
for(fun in list(min, max)){
      cell <- cpr %>% filter(pd_obs == fun(pd_obs)) %>% rownames()
      spp <- colnames(ds$comm)[ds$comm[cell,] == 1]
      phyplot(ds$tree, connect = spp, 
              type = "fan", show.tip.label = F,
              main = paste("taxa found in", cell))
}
```

#### Manual approach

Canaper and phyloregion are pretty great, but how are these spatial biodiversity metrics actually being calculated under the hood? Let's reproduce some of these results with a slightly more manual approach.

```{r manual}

# add clades to community matrix
pcomm <- build_clade_ranges(ds$tree, ds$comm) # from functions.r
# dim(ds$comm)
# dim(pcomm)

# clade-level metrics
range_size <- colSums(pcomm)
branch_length <- ds$tree$edge.length / sum(ds$tree$edge.length)
phyplot(ds$tree, value = branch_length,
        show.tip.label = F, type = "fan", main = "branch length")
phyplot(ds$tree, value = log(1 / range_size),
        show.tip.label = F, type = "fan", main = "log endemism")

pd <- pcomm %>%
      apply(1, function(x) x * branch_length) %>%
      colSums()
ds$xy %>% mutate(pd = pd) %>% carto("pd")

pe <- pcomm %>%
      apply(1, function(x) x * branch_length / range_size) %>%
      colSums()
ds$xy %>% mutate(pe = pe) %>% carto("pe")

```

#### Exercises:

-   Compute these metrics with one of the other datasets.
-   Species richness can be considered PD on a star phylogeny. Prove it.
-   Manual calculations make it clear that these diversity metrics are simply weighted summaries of the `pcomm` matrix, which invites any number of additional weighting schemes. Make a version weighted by: taxon endangerment; site endangerment; or taxon-site habitat importance (using fake simulated data for these variables).

```{r alpha_ex}

```

#### Significance testing

To get a sense for how noteworthy an observed high or low diversity metric is, it can be helpful to compare the observed value to a null distribution. Let's demonstrate this using the acacia dataset. The `cpr_rand_test()` function randomly permutes the community matrix many times, calculating the diversity of each location each time to derive a distribution of null diversity values for each location.

CANAPE (Mishler et al. 2014) is a special application of this randomization approach, used to identify areas of neo-endemism and paleo-endemism.

```{r sig}
ds <- acacia

cpr <- cpr_rand_test(ds$comm, ds$tree, 
                     null_model = "curveball",
                     n_reps = 101, n_iterations = 10000) %>%
      bind_cols(ds$xy)
# glimpse(cpr)

# raw PD patterns here are very different from randomized quantiles
cpr %>% carto("pd_obs")
cpr %>% carto("pd_obs_p_upper")cpr %>%
      ggplot(aes(pd_obs, pd_obs_p_upper)) +
      geom_point()

# CANAPE
cpr %>%
      cpr_classify_endem() %>%
      carto("endem_type") +
      scale_fill_manual(
            values = c("mediumpurple1", "red", "gray80", "blue", "darkorchid4"),
            breaks = c("mixed", "neo", "not significant", "paleo", "super"))

```

Null models can be constructed in many different ways, and these differences can have a major effect on results. We won't get too far into the weeds here, but let's illustrate how a few different null models affect PD significance values for this dataset:

```{r nulls}

# ?vegan::commsim
curveball <- cpr %>% 
      mutate(model = "curveball")
r0 <- cpr_rand_test(ds$comm, ds$tree, null_model = "r0", metrics = "pd", 
                    n_reps = 101) %>%
      bind_cols(ds$xy) %>%
      mutate(model = "r0")
c0 <- cpr_rand_test(ds$comm, ds$tree, null_model = "c0", metrics = "pd", 
                    n_reps = 101) %>%
      bind_cols(ds$xy) %>%
      mutate(model = "c0")

bind_rows(curveball, r0, c0) %>%
      mutate(pd_sig = case_when(pd_obs_p_upper > .95 ~ "high",
                                pd_obs_p_upper < .05 ~ "low",
                                TRUE ~ "no")) %>%
      carto("pd_sig") + 
      facet_wrap(~model, nrow = 1) +
      scale_fill_manual(values = c("red", "blue", "gray"))

```

Before moving on from randomizations, it's important to note that the size of the geographic region in an analysis can strongly influence the results, because it alters the species pool to which a site is compared during randomization. Let's look at three different extents:

```{r scale}

# a function to subset the sites in a dataset
filter_sites <- function(ds, selection){
      dsf <- ds
      dsf$comm <- dsf$comm[selection, ]
      dsf$xy <- dsf$xy[selection, ]
      dsf<- clean_dataset(dsf)
      return(dsf)
}

# filter mammals data to three nested regions of different sizes
americas <- mammals %>% 
      filter_sites(mammals$xy$x < -34)
s_america <- mammals %>% 
      filter_sites(between(mammals$xy$x, -82, -34) 
                   & mammals$xy$y < 13)
amazon <- mammals %>% 
      filter_sites(between(mammals$xy$x, -75, -50) 
                   & between(mammals$xy$y, -15, 5))

# compute pd significance for each dataset
cpr_americas <- cpr_rand_test(
      americas$comm, americas$tree, 
      null_model = "curveball", metrics = "pd", n_reps = 21) %>%
      bind_cols(americas$xy) %>%
      mutate(region = "americas")
cpr_s_america <- cpr_rand_test(
      s_america$comm, s_america$tree, 
      null_model = "curveball", metrics = "pd", n_reps = 21) %>%
      bind_cols(s_america$xy) %>%
      mutate(region = "south america")
cpr_amazon <- cpr_rand_test(
      amazon$comm, amazon$tree, 
      null_model = "curveball", metrics = "pd", n_reps = 21) %>%
      bind_cols(amazon$xy) %>%
      mutate(region = "amazon")
cpr <- bind_rows(cpr_americas, cpr_s_america, cpr_amazon) %>%
      mutate(region = factor(region, levels = unique(region))) %>%
      mutate(pd_sig = case_when(pd_obs_p_upper > .95 ~ "high",
                                pd_obs_p_upper < .05 ~ "low",
                                TRUE ~ "no"))

# maps
cpr %>%
      carto("pd_sig") +
      facet_wrap(~ region, nrow = 1) +
      scale_fill_manual(values = c("red", "blue", "gray"))
cpr %>%
      filter(x %in% x[region == "amazon"],
             y %in% y[region == "amazon"]) %>%
      ggplot(aes(x, y, fill = pd_sig)) +
      geom_raster() +
      facet_wrap(~ region, nrow = 1) +
      theme_void() + theme(legend.position = "top") + coord_fixed() +
      scale_fill_manual(values = c("red", "blue", "gray"))
```

#### Facets of phylogenetic diversity

Phylogenetic branch lengths can represent different macroevolutionary variables, each with different implications for biogeography and conservation. We've indirectly met some different "phylogenetic diversity facets" already, through RPD (which is computed in `canaper`). Let's quickly explore them in a bit more detail.

```{r facets}

ds <- flora

scale_edges <- function(tree){
      tree$edge.length <- tree$edge.length / sum(tree$edge.length)
      return(tree)
}

# load trees
chronogram <- read.nexus("../../data/thornhill_2017/thornhill_2017_chronogram.nex") %>%
      scale_edges()
phylogram <- read.nexus("../../data/thornhill_2017/thornhill_2017_phylogram.nex") %>%
      scale_edges()

# make cladogram
cladogram <- phylogram
cladogram$edge.length <- rep(1, length(cladogram$edge.length))
cladogram <- scale_edges(cladogram)

# plot trees
par(mfrow = c(1, 3))
phyplot(chronogram, chronogram$edge.length,
        type = "fan", show.tip.label = F, main = "time")
phyplot(phylogram, phylogram$edge.length,
        type = "fan", show.tip.label = F, main = "divergence")
phyplot(cladogram, cladogram$edge.length,
        type = "fan", show.tip.label = F, main = "cladogenesis")
# phyplot(chronogram, phylogram$edge.length / chronogram$edge.length,
#                   type = "fan", show.tip.label = F, main = "dynamism")
dev.off()


# format as sparse matrix for phyloregion package
scomm <- ds$comm %>% Matrix(sparse = TRUE)

div <- ds$xy %>%
      mutate(divergence = PD(scomm, phylogram),
             time = PD(scomm, chronogram),
             diversification = PD(scomm, cladogram))
div %>%
      pivot_longer(c(divergence, time, diversification), 
                   "facet", "value") %>%
      group_by(facet) %>%
      mutate(rank = rank(value)) %>%
      carto("rank") + 
      facet_wrap(~facet, nrow = 1)

div %>%
      mutate(rpd = time / diversification) %>% 
      carto("rpd")

div %>%
      mutate(dynamism = divergence / time) %>%
      carto("dynamism")

```


#### Exercises:

-   Verify that `rpd` as returned by the `cpr_rand_test()` function using a chronogram is identical to the ratio of chronogram PD to cladogram PD calculated immediately above.
-   Make a map of the ratio of phylogram PD to cladogram PD, which measures the relative contribution of anagenesis versus cladogenesis to a community's history. Where is this highest and lowest?

```{r alpha_ex2}

```


------------------------------------------------------------------------

# BETA DIVERSITY

Analyses of beta diversity focus on compositional differences among local communities. Phylogenetic versions of these analyses represent differences between communities as the difference in the portion of the phylogenetic tree that's found in each site.

#### Turnover

We'll begin by creating a pairwise turnover matrix with a row and column for each site. We can compare this to species-based turnover and geographic distance. Further exploration of these turnover patterns could use a modeling approach like GDM (Ferrier 2007) to understand how factors like environmental differences explain phylogenetic turnover, but we will not cover that here.

```{r beta}

ds <- mammals

# construct dissimlarity matrices
phy_beta <- ds$comm %>% # phylo sorensen's
      Matrix(sparse = TRUE) %>%
      phylobeta(ds$tree) %>%
      pluck("phylo.beta.sor") 
sp_beta <- ds$comm %>% # species sorensen's
      betapart::beta.pair() %>%
      pluck("beta.sor")
geo_dist <- ds$xy %>% # geographic distance
      geosphere::distm()

# plot geographic distance against phylogenetic turnover
ss <- sample(length(geo_dist), 10000)
tibble(distance = geo_dist[ss],
       phylo_sor = as.matrix(phy_beta)[ss],
       species_sor = as.matrix(sp_beta)[ss]) %>%
      pivot_longer(-distance, "stat", "value") %>%
      ggplot(aes(distance, value, color = stat)) +
      geom_point(size = .5) +
      geom_smooth(aes(group = stat), color = "black")
```

#### Regionalization

We can also use turnover data to make maps of phylogenetic similarity. Let's look at two approaches: a discrete classification of phylogenetic regions using a cluster analysis from the `phyloregions` package, and a continuous ordination map that plots similar sites in similar colors:

```{r regions}

# phylogenetic regionalization (spatial cluster analysis)
k <- 25
phyloregion(phy_beta, k = k) %>%
      pluck("membership") %>%
      bind_cols(ds$xy) %>%
      mutate(cluster = factor(cluster))%>%
      carto("cluster") +
      scale_fill_manual(values = unname(pals::alphabet(k)))

# RGB visualization
# ord <- vegan::metaMDS(phy_beta, k = 3, trymax = 50)$points
# saveRDS(ord$points, "../../data/cache/mammals_ord.rds")
ord <- readRDS("../../data/cache/mammals_ord.rds")
ord %>%
      as.data.frame() %>%
      mutate_all(function(x) rank(x)/max(rank(x))) %>%
      bind_cols(ds$xy) %>%
      mutate(color = rgb(MDS1, MDS2, MDS3)) %>%
      carto("color") +
      scale_fill_identity()
```

#### Exercises:

- The above demonstration used Sorensen's distance as a measure of phylogenetic community difference. Try things with Jaccard's distance. How do the results differ?
- Experiment with different numbers of clusters and different cluster methods in the `phyloregion()` function, to explore how robust regional boundaries are to these parameter choices.

```{r beta_ex}

```

------------------------------------------------------------------------

# CONSERVATION PRIORITIZATION

Phylogenetic diversity is useful for basic research, but from the beginning it has also been proposed as an applied conservation tool. Let's use it to run a conservation prioritization analysis. We'll focus on the California vascular flora dataset for this exercise, which has been used for conservation prioritization in the past ([Kling et al. 2018](https://doi.org/10.1098/rstb.2017.0397)).

Optimal reserve design is a major computational challenge, due to the extremely large number of sets of sites that could comprise a protected area network. Our computational workhorse will be the `prioritizr` library, a very flexible and powerful conservation planning toolkit that uses linear solvers to identify optimal conservation solutions.

First we need to get our data formatted as `raster` objects. We'll also load an additional data layer of current protected areas, derived from Kling et al. (2018).

```{r prioritize_setup}

ds <- flora

# conservation features
features <- ds$xy %>%
      bind_cols(ds$comm) %>%
      rasterFromXYZ()

# spatial planning units
units <- features[[1]] %>% 
      reclassify(c(-Inf, Inf, 1))

# current protected areas
protected <- raster("../../data/thornhill_2017/kling_2018_protection_status.tif")
# plot(protected)
protected <- protected > .66
prot_area <- sum(protected[], na.rm = T)

# harmonize spatial coverage
overlap <- protected + units
protected <- protected %>% crop(overlap) %>% mask(overlap)
units <- units %>% crop(overlap) %>% mask(overlap)
features <- features %>% crop(overlap) %>% mask(overlap)
crs(protected) <- crs(units) <- crs(features)

```

The `prioritizr` library has some native functionality for incorporating phylogenies into conservation prioritization, so let's start there.

```{r prioritize}

# compare existing protected area network to 
# an optimal network of the same size
prob <- problem(units, features) %>%
      add_max_phylo_div_objective(budget = prot_area,
                                  tree = ds$tree) %>%
      add_relative_targets(0.1) %>%
      add_binary_decisions() %>%
      add_rsymphony_solver(gap = 0)
soln <- solve(prob)
stack(soln, protected) %>% 
      setNames(c("optimal", "existing")) %>%
      plot(col = c("gray", "orange"))

# identify the highest-priority sites to add to existing protected areas
prob <- prob %>%
      add_max_phylo_div_objective(budget = prot_area + 10,
                                  tree = ds$tree) %>%
      add_locked_in_constraints(locked_in = protected)
soln <- solve(prob)
plot(soln + protected, 
     col = c("gray", "orange", "darkgreen"))

```

While it incorporates phylogeny, this method is still "tip-centric" in the sense that its objective is to maximize the phylogenetic diversity of the tips protected across the reserve network. An alternative that's arguably more consistent with other spatial phylogenetic metrics is to build out a phylogenetic community matrix, and weight each taxon by its branch length during prioritization. We can see this yields a somewhat different result:

```{r prioritize_pcomm}

pcomm <- build_clade_ranges(ds$tree, ds$comm)

pfeatures <- ds$xy %>%
      bind_cols(pcomm) %>%
      rasterFromXYZ()

# define and solve prioritization problem
prob <- problem(units, pfeatures) %>%
      add_max_features_objective(budget = prot_area + 10) %>%
      add_feature_weights(weights = ds$tree$edge.length) %>%
      add_relative_targets(0.1) %>%
      add_binary_decisions() %>%
      add_locked_in_constraints(locked_in = protected) %>%
      add_rsymphony_solver(gap = 0)
soln <- solve(prob)
plot(soln + protected, 
     col = c("gray", "orange", "darkgreen"))
```

#### Exercises:

- These exercises were based on the chronogram, which we packaged up with the California flora dataset. Replicate them with the phylogram and cladogram. Do the conservation priorities change?
- `prioritizr` has functionality to consider many additional factors, such as land cost, reserve shape, and reserve connectivity. A simple example is `add_neighbor_constraints(k = 2)`, which would force proposed reserves to have at least two neighbors. Like other constraints this will likely add a lot of computational time, so try it when you have a few minutes to wait.

```{r prioritize_ex}

```

------------------------------------------------------------------------

And that is all!
